EBASE manuscript

### 2023-01-08 manu-draft.qmd
adding figures for apasumdat and apasumdatmean, results text for apasumdat

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.  A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.  Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Additional functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.  This differs from the BASEmetab approach where the optimization period is set at one day.  All results provided herein were generated using the EBASE R package.

@@

### EBASE comparison with modelled observations

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with FWOXY. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-simapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).  The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251.

Several configurations of the EBASE model can influence how the metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series. First, parameter estimates were compared to the known values after changing characteristics of their prior distributions.  Second, parameter estimates were compared using a model period of optimization for one day and one week of observations.  Details of these analyses are as follows.

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations.  The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors. For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.  Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.  Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior.  The mean of the $b$ prior was varied as 50%, 100%, and 200% (the upper limit used in the software) of the default and the standard deviation was varied as before.  This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).  Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.  For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for  24 hour time period if the sampling interval is 15 minutes).  Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.  As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.  An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 unique analyses for the assessment of changing standard deviations of all parameters and 18 unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.

Results for the sensitivity analyses are shown in ......

@@

>>>>

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.  A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.  Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series. This differs from the BASEmetab approach where the optimization period is set at one day. Sensitivity of the EBASE method to different optimization periods is described in @sec-cmpmod. All results provided herein were generated using the EBASE R package.

@@

### EBASE comparison with modelled observations {#sec-cmpmod}

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with FWOXY. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay ([@fig-simapa]a), such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange ([@fig-simapa]b and c).  The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251 $\buni$.

Several configurations of the EBASE model can influence how the parameters and metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the results returned by EBASE, where in all cases, the results were compared to the known values in the simulated time series. First, results were compared to the known values after changing characteristics of their prior distributions.  Second, results were compared using a model period of optimization for one day and one week of observations.  Details of these analyses are as follows.

The standard prior distributions for the EBASE parameters follow normal distributions with defined means and standard deviations (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]).  The sensitivity analysis evaluated changes in the distributions by varying the standard deviations from small to large to test the effect of more or less constrained ranges, respectively, on the priors. For each parameter, low, medium, and high values for the standard deviation were assessed as 10%, 100%, and 1000% of the default values.  Each unique combination of low, medium, and high values for the standard deviation of each parameter were evaluated, creating 3$^3$ = 27 different combinations of prior distributions that were used with the simulated time series in EBASE.  Because initial assessments indicated that all parameter estimates were especially sensitive to changes in the prior distribution for the $b$ parameter, a second sensitivity analysis evaluated additional changes in both the mean and standard deviation of the prior.  The mean of the $b$ prior was varied as 50%, 100%, and 200% (the upper limit used in the software) of the default and the standard deviation was varied as before.  This resulted in 3$^2$ = 9 unique combinations of means and standard deviations for the $b$ prior that were evaluated.

The optimization period used for a time series determines how much data are used to estimate the three EBASE parameters ($a$, $r$, and $b$).  Unlike BASEmetab, the EBASE software allows flexibility in the period of time used for optimization, with a minimum period of time being one day and a maximum period of being the length of the time series.  For a chosen optimization period, a single estimate for $r$ and $b$ is returned by the model, whereas $n$ number of observations equal to the number of time steps in each optimization period are returned for the $a$ parameter (e.g., 96 values for a 24 hour time period if the sampling interval is 15 minutes).  Short periods of observation may return unstable parameter estimates if forcing factors (e.g., wind, temperature) that affect estuarine biogeochemical rates occur at time scales longer the chosen period of optimization for EBASE.  As such, potentially longer optimization periods may return more stable and robust estimates given that more data are used for parameter estimation.  An optimization period of one day was compared to results using an optimization period of seven days for the simulated one-year time series. Each optimization period was evaluated for every unique combination of prior distributions, totaling 54 (27 times the two optimization periods) unique analyses for the assessment of changing standard deviations of all parameters and 18 (9 times the two optimization periods) unique analyses for the assessment of changing the mean and standard deviation of only the $b$ parameter.

Changing the variance of each of the prior distributions affected the ability of EBASE to return the known parameters from the simulated time series. @fig-apasumdat shows a summary of the comparisons of the simulated data with EBASE for each unique combination of prior distributions and two optimization periods of one and seven days.  The results are compared using R-squared ($R^2$) values from the linear fit between the simulated data and results from EBASE.  Darker colors represent lower $R^2$ values and can be used to assess how changing the variance of the priors affects the ability of EBASE to estimate the true parameter.

Several conclusions can be made from the results in @fig-apasumdat based on the patterns of the $R^2$ values. Overall, the model was able to estimate the $a$ parameter with reasonable precision regardless of any of the prior distributions, whereas most of the variation was observed in the $r$ and $b$ parameters.  Changing the variance of the $a$ parameter had little effect on any of the comparisons.  Smaller variance of the prior distribution for the $r$ parameter generally contributed to lower $R^2$ for all parameters, with the precision being worse for the 7-day optimization period (e.g., the $a$ parameter).  Larger variance of the prior distribution for the $b$ parameter generally contributed to lower $R^2$ only for the gas exchange estimate (D), although this effect was minimized with the 7-day optimization period.  As such, the results suggest that the default prior distributions used in EBASE may be appropriate in some cases, but increasing the prior variance for $a$ and $r$ and decreasing the variance for $b$ will likely produce more optimal solutions.

@@

![Comparison of EBASE results with known metabolic parameters from a simulated time series. Results are compared for unique combination of prior distributions, where the variances are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.  The $b$ parameter is not evaluated because it is a fixed constant for the simulated data.](figs/apasumdat.png){#fig-apasumdat}

![Comparison of EBASE results with known metabolic parameters from a simulated time series. Results are compared for unique combination of prior distributions, where the variances and means for the prior of the $b$ parameter are evaluated from low (L), medium (M), to high (H) for a one-day and seven-day model optimization period. The metabolic results for dissolved oxygen (DO), production (P), respiration (R), gas exchange (D), and the $a$ parameter are evaluated based on the $R^2$ value of the linear fit between the EBASE results and simulated values.](figs/apasumdatmean.png){#fig-apasumdatmean}
